---
title: "Sublethal"
author: "Layla"
date: "2026-03-25"
output:
  word_document: default
  html_document: default
  pdf_document: default
---

## GRAPH 1: Total reproduction per individual vs treatment

```{r}

data <- read.csv("C:/Users/layla/OneDrive/Desktop/Sublethal/data/Sub Lethal Data.csv")
library(dplyr)
library(ggplot2)
library(tidyverse)
colnames(data) <- trimws(colnames(data))
colnames(data)
```



```{r}
data <- read.csv("C:/Users/layla/OneDrive/Desktop/Sublethal/data/Sub Lethal Data.csv")
library(dplyr)
library(ggplot2)
library(tidyverse)

colnames(data) <- trimws(colnames(data))

data$Treatment <- trimws(data$Treatment)
data$Treatment <- gsub("\\s+", " ", data$Treatment)

unique(data$Treatment)

data$Treatment <- recode(data$Treatment,
                         "0 ppm" = "0",
                         "1 ppm" = "1",
                         "2 ppm" = "2",
                         "10 ppm" = "10")

data$Treatment <- factor(data$Treatment, levels = c("0", "1", "2", "10"))

unique(data$Treatment)

nauplii_cols <- c("Nauplii..1.", "Nauplii..2.", "Nauplii..3.", "Nauplii..4.", "Nauplii..5.")

data[nauplii_cols] <- lapply(data[nauplii_cols], as.numeric)
data[nauplii_cols][is.na(data[nauplii_cols])] <- 0

individual_totals <- data %>%
  group_by(Individual.ID, Treatment) %>%
  summarise(
    clutch1 = first(Nauplii..1.[Nauplii..1. > 0], default = 0),
    clutch2 = first(Nauplii..2.[Nauplii..2. > 0], default = 0),
    clutch3 = first(Nauplii..3.[Nauplii..3. > 0], default = 0),
    clutch4 = first(Nauplii..4.[Nauplii..4. > 0], default = 0),
    clutch5 = first(Nauplii..5.[Nauplii..5. > 0], default = 0),
    total_nauplii = clutch1 + clutch2 + clutch3 + clutch4 + clutch5,
    .groups = "drop"
  )

ggplot(individual_totals, aes(x = Treatment, y = total_nauplii)) +
  geom_jitter(width = 0.2) +
  labs(
    title = "Total Reproduction per Individual vs Treatment",
    x = "Treatment (ppm)",
    y = "Total Nauplii per Individual"
  )
```
```{r}
ggplot(individual_totals, aes(x = Treatment, y = total_nauplii)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  labs(
    title = "Total Reproduction per Individual vs Treatment",
    x = "Treatment (ppm)",
    y = "Total Nauplii per Individual"
  )
```
```{r}
kruskal.test(total_nauplii ~ Treatment, data = individual_totals)

pairwise.wilcox.test(individual_totals$total_nauplii,
                     individual_totals$Treatment,
                     p.adjust.method = "bonferroni")
```


## GRAPH 2: Clutch success / number of clutches per individual

```{r}

individual_totals <- individual_totals %>%
  mutate(
    num_clutches = (clutch1 > 0) + (clutch2 > 0) + (clutch3 > 0) + (clutch4 > 0) + (clutch5 > 0)
  )

ggplot(individual_totals, aes(x = factor(Treatment), y = num_clutches)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  labs(
    title = "Number of Clutches per Individual vs Treatment",
    x = "Treatment (ppm)",
    y = "Number of Clutches per Individual"
  )

```
```{r}
individual_totals$Treatment <- factor(individual_totals$Treatment,
                                      levels = c(0, 1, 2, 10))

kruskal.test(num_clutches ~ Treatment, data = individual_totals)
pairwise.wilcox.test(individual_totals$num_clutches,
                     individual_totals$Treatment,
                     p.adjust.method = "bonferroni",
                     exact = FALSE)
```
## GRAPH 3: Nauplii per clutch (clutch size)

```{r}
clutch_long <- individual_totals %>%
  pivot_longer(
    cols = c(clutch1, clutch2, clutch3, clutch4, clutch5),
    names_to = "clutch_number",
    values_to = "nauplii"
  ) %>%
  filter(nauplii > 0)

clutch_long$Treatment <- factor(clutch_long$Treatment,
                                levels = c(0, 1, 2, 10))

ggplot(clutch_long, aes(x = Treatment, y = nauplii)) +
  geom_boxplot(alpha = 0.5) +
  labs(
    title = "Nauplii per Clutch vs Treatment",
    x = "Treatment (ppm)",
    y = "Nauplii per Clutch"
  )
```
```{r}
kruskal.test(nauplii ~ Treatment, data = clutch_long)

# general linear model 
# negative binomial 
# poisson 
# glmmTMB

```
```{r}
pairwise.wilcox.test(clutch_long$nauplii,
                     clutch_long$Treatment,
                     p.adjust.method = "bonferroni",
                     exact = FALSE)
```

## GRAPH 4: Time to reproduction (timing)
### ( send to slack with david)

```{r}
data$Date <- as.Date(data$Date, format = "%m/%d/%Y")
data$Date.of.Transfer <- as.Date(data$Date.of.Transfer, format = "%m/%d/%Y")

clutch_time <- data %>%
  pivot_longer(
    cols = c("Nauplii..1.", "Nauplii..2.", "Nauplii..3.", "Nauplii..4.", "Nauplii..5."),
    names_to = "clutch_number",
    values_to = "nauplii"
  ) %>%
  mutate(
    nauplii = as.numeric(nauplii),
    clutch_num = as.numeric(gsub("Nauplii\\.\\.|\\.", "", clutch_number))
  ) %>%
  filter(nauplii > 0) %>%
  group_by(Individual.ID, Treatment, clutch_num) %>%
  summarise(
    first_date = min(Date),
    start_date = min(Date.of.Transfer),
    .groups = "drop"
  ) %>%
  arrange(Individual.ID, clutch_num) %>%
  group_by(Individual.ID) %>%
  mutate(
    interval_days = ifelse(
      clutch_num == 1,
      as.numeric(first_date - start_date),
      as.numeric(first_date - lag(first_date))
    ),
    interval_label = case_when(
      clutch_num == 1 ~ "Transfer to clutch 1",
      clutch_num == 2 ~ "Clutch 1 to 2",
      clutch_num == 3 ~ "Clutch 2 to 3",
      clutch_num == 4 ~ "Clutch 3 to 4",
      clutch_num == 5 ~ "Clutch 4 to 5"
    )
  ) %>%
  ungroup()

ggplot(clutch_time, aes(x = interval_label, y = interval_days, color = Treatment)) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(width = 0.75)) +
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.75)) +
  labs(
    title = "Timing of Reproduction Across Clutches",
    x = "Reproductive interval",
    y = "Days",
    color = "Treatment (ppm)"
  )
```
```{r}
ggplot(clutch_time, aes(x = Treatment, y = interval_days)) +
  geom_boxplot() +
  geom_jitter(width = 0.2) +
  facet_wrap(~ interval_label) +
  labs(
    title = "Reproductive Timing by Treatment",
    x = "Treatment (ppm)",
    y = "Days"
  )
```
```{r}

table(clutch_time$interval_label, clutch_time$Treatment)

# Transfer to first clutch
kruskal.test(interval_days ~ Treatment,
             data = subset(clutch_time, interval_label == "Transfer to clutch 1"))

# Clutch 1 to 2
kruskal.test(interval_days ~ Treatment,
             data = subset(clutch_time, interval_label == "Clutch 1 to 2"))

# Clutch 2 to 3
kruskal.test(interval_days ~ Treatment,
             data = subset(clutch_time, interval_label == "Clutch 2 to 3"))

# Clutch 3 to 4
kruskal.test(interval_days ~ Treatment,
             data = subset(clutch_time, interval_label == "Clutch 3 to 4"))
```
```{r}
pairwise.wilcox.test(
  subset(clutch_time, interval_label == "Clutch 3 to 4")$interval_days,
  subset(clutch_time, interval_label == "Clutch 3 to 4")$Treatment,
  p.adjust.method = "bonferroni",
  exact = FALSE
)
```

## GRAPH 5: Cumulative reproduction over time ( run this as a GLM)

```{r}
# make sure dates are correct
data$Date <- as.Date(data$Date, format = "%m/%d/%Y")
data$Date.of.Transfer <- as.Date(data$Date.of.Transfer, format = "%m/%d/%Y")

# reshape nauplii data
cum_data <- data %>%
  pivot_longer(
    cols = c("Nauplii..1.", "Nauplii..2.", "Nauplii..3.", "Nauplii..4.", "Nauplii..5."),
    names_to = "clutch",
    values_to = "nauplii"
  ) %>%
  mutate(nauplii = as.numeric(nauplii)) %>%
  filter(nauplii > 0)

# keep only FIRST appearance of each clutch
cum_data <- cum_data %>%
  group_by(Individual.ID, Treatment, clutch) %>%
  summarise(
    clutch_date = min(Date),
    transfer_date = min(Date.of.Transfer),
    nauplii = first(nauplii),
    .groups = "drop"
  )

# calculate days since transfer
cum_data <- cum_data %>%
  mutate(days_since = as.numeric(clutch_date - transfer_date))

# order data and calculate cumulative reproduction per individual
cum_data <- cum_data %>%
  arrange(Individual.ID, days_since) %>%
  group_by(Individual.ID) %>%
  mutate(cum_nauplii = cumsum(nauplii)) %>%
  ungroup()

cum_summary <- cum_data %>%
  group_by(Treatment, days_since) %>%
  summarise(mean_cum = mean(cum_nauplii), .groups = "drop")

ggplot(cum_summary, aes(x = days_since, y = mean_cum, color = Treatment, group = Treatment)) +
  geom_line(size = 1) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Cumulative Reproduction Over Time",
    x = "Days Since Transfer",
    y = "Mean Cumulative Nauplii",
    color = "Treatment (ppm)"
  )
```

## GRAPH 6: Clutch sequence plot
```{r}
clutch_mean <- clutch_long %>%
  group_by(Treatment, clutch_number) %>%
  summarise(mean_nauplii = mean(nauplii), .groups = "drop")

clutch_mean$clutch_number <- factor(clutch_mean$clutch_number,
                                    levels = c("clutch1", "clutch2", "clutch3", "clutch4", "clutch5"),
                                    labels = c("1", "2", "3", "4", "5"))

ggplot(clutch_mean, aes(x = clutch_number, y = mean_nauplii, color = Treatment, group = Treatment)) +
  geom_line() +
  geom_point() +
  labs(
    title = "Clutch Sequence Plot",
    x = "Clutch Number",
    y = "Mean Nauplii per Clutch",
    color = "Treatment (ppm)"
  )
```
```{r}
clutch_long$clutch_number <- factor(clutch_long$clutch_number,
                                    levels = c("clutch1", "clutch2", "clutch3", "clutch4", "clutch5"),
                                    labels = c("1", "2", "3", "4", "5"))

ggplot(clutch_long, aes(x = clutch_number, y = nauplii, group = Individual.ID, color = Treatment)) +
  geom_line(alpha = 0.4) +
  geom_point() +
  labs(
    title = "Clutch Sequence Plot by Individual",
    x = "Clutch Number",
    y = "Nauplii per Clutch",
    color = "Treatment (ppm)"
  )

```

## Running the model 

```{r}
names(data)
head(data)
data$Date <- as.Date(data$Date, format = "%m/%d/%Y")
data$Death.Date..if.before.21.days. <- as.Date(
  data$Death.Date..if.before.21.days.,
  format = "%m/%d/%Y"
)

# Build the offspring columns
nauplii_cols <- c("Nauplii..1.", "Nauplii..2.", "Nauplii..3.", "Nauplii..4.", "Nauplii..5.")

data[nauplii_cols] <- lapply(data[nauplii_cols], as.numeric)

data$cum_offspring <- rowSums(data[nauplii_cols], na.rm = TRUE)

# Sort by individual and date
data <- data[order(data$Individual.ID, data$Date), ]

#Create offspring produced at each observation
data$all_egg <- ave(
  data$cum_offspring,
  data$Individual.ID,
  FUN = function(x) c(x[1], diff(x))
)

data$all_egg[is.na(data$all_egg)] <- 0
data$all_egg[data$all_egg < 0] <- 0

# Create age in days
data$week <- ave(
  as.numeric(data$Date),
  data$Individual.ID,
  FUN = function(x) x - min(x)
)

# Create alive/dead status
data$Alive <- ifelse(
  !is.na(data$Death.Date..if.before.21.days.) &
    data$Date >= data$Death.Date..if.before.21.days.,
  0,
  1
)

#Create the ID column the model expects
data$snail_id <- data$Individual.ID

#Keep only the needed columns
life_table_data <- data[, c("week", "snail_id", "all_egg", "Alive", "Treatment")]

head(life_table_data)
str(life_table_data)

#Clean treatment values
life_table_data$Treatment <- trimws(life_table_data$Treatment)
life_table_data$Treatment <- gsub("\\s+", " ", life_table_data$Treatment)

unique(life_table_data$Treatment)

#split by treatment
treat_0 <- subset(life_table_data, Treatment == 0)
treat_1 <- subset(life_table_data, Treatment == 1)
treat_2 <- subset(life_table_data, Treatment == 2)
treat_10 <- subset(life_table_data, Treatment == 10)

```

- Issue  the date of the experiment is not he same thing as the age of the orgaizm, so for the day columb we need the age of the animel ( find that in phase 1-2)

- s there evidance that there are individuals with egg sacks and never the naplei after ( calukate the perportion of lost clutches)

- trading off current reproduction for future self preservation and survival 

- Three columbs, ID, #wihc clutch, sussess or failure (where nap present)

```{r}
# Run the little-r model

Euler <- function(r, lx_times_mx, time){
  sum(lx_times_mx * exp(-r * time)) - 1
}

little.r <- function(treatment_dataset){
  lx_and_week <- aggregate(Alive ~ week, FUN = mean, data = treatment_dataset, drop = FALSE)
  time <- as.numeric(lx_and_week[, 1])
  lx <- as.numeric(lx_and_week[, 2])

  mx <- as.numeric(aggregate(all_egg ~ week, FUN = mean, data = treatment_dataset, drop = FALSE)[, 2])
  mx[!is.finite(mx)] <- 0

  lx_times_mx <- lx * mx

  uniroot(Euler,
          interval = c(-10, 10),
          lx_times_mx = lx_times_mx,
          time = time)$root
}

little.r(treat_0)
little.r(treat_1)
little.r(treat_2)
little.r(treat_10)
```
```{r}
head(life_table_data)
summary(life_table_data)
table(life_table_data$Alive, useNA = "ifany")
```
The intrinsic rate of increase (r) declined with increasing treatment concentration. Control (0 ppm) and low exposure (1 ppm) treatments showed similar population growth rates (~2.67), while moderate exposure (2 ppm) caused a slight reduction (2.55). The highest treatment (10 ppm) resulted in a substantial decrease in population growth (1.76), indicating strong negative effects on survival and/or reproduction.

```{r}
hist(little.r.CIs(treat_0, 1000))
hist(little.r.CIs(treat_10, 1000))
quantile(little.r.CIs(treat_0, 1000), c(0.025, 0.975))
```
```{r}
quantile(little.r.CIs(treat_0, 1000), c(0.025, 0.975))
quantile(little.r.CIs(treat_1, 1000), c(0.025, 0.975))
quantile(little.r.CIs(treat_2, 1000), c(0.025, 0.975))
quantile(little.r.CIs(treat_10, 1000), c(0.025, 0.975))
```
```{r}
results <- data.frame(
  Treatment = c("0 ppm", "1 ppm", "2 ppm", "10 ppm"),
  lower = c(1.816, 2.009, 1.249, 1.266),
  upper = c(3.190, 3.091, 3.159, 2.212),
  mean = c(2.674, 2.675, 2.545, 1.760)
)

plot(1:4, results$mean,
     ylim = c(1, 3.5),
     xaxt = "n",
     xlab = "Treatment",
     ylab = "r")

axis(1, at = 1:4, labels = results$Treatment)

arrows(1:4, results$lower, 1:4, results$upper, angle = 90, code = 3, length = 0.05)
```


